Method to infer relatedness using low coverage ancient DNA

### 2022-03-14 paper_overleaf/main.tex

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{csvsimple}
\usepackage[section]{placeins}

@@

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BH}{\mathbf{H}}
\newcommand{\Btheta}{\pmb{\theta}}

@@

\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness upto 3rd degree, while additionally differentiating between siblings and parent-child. We have developed another HMM to estimate ROH proportions, that can be utilized by the relatedness estimation model to output better estimates. 
\end{abstract}

@@

Identifying related individuals is a common task in applications of genetics. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify missing persons following a disaster \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential preprocessing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_kinship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

@@

Commonly, pairs of relateded individuals are  identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_kinship_2020}.  One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999,mcpeek_statistical_2000}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize a pair of individuals as related by a certain degree and nature of relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings with and without ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On left, the parents have unrelated chromosomes, and on right one parent has homozygous pair of chromosomes. Bottom row shows the expected proportion of differences along the chromosome in both the cases.}
    \label{fig0:schematic}
\end{figure}

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in  diploid genotype data, estimates IBD probabilities from average of observed IBS, and using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods solve this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}. These methods account for the uncertainity in genotype calls by summing over all possible genotypes, weighted by the genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional problems like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage to get a good estimate of genotype likelihoods\cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. In many cases, ancient DNA is captured with a SNP array, and the SNPs are ascertained to variable sites determined from certain individuals. Relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust kinship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

READ is a software that does not require population allele frequencies or a reference panel, and can work with around 0.5 x coverage in presence of ascertainment bias. The software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudohaploids) are then compared to other individuals to calculate pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify upto only second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIn (Kinship Inference), a Hidden Markov Model (HMM) to estimate relatedness from low-coverage ancient DNA data. KIn can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIn is also able to incorporate ROH, contamination and ascertainment bias. We validate the performance of KIn using simulations and show that we are able to infer relatedness in a neandertal data set, and a Bronze age dataset.

@@

The core of  KIn, is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts in each sample, and contamination estimates into account. If the locations of ROH tracts are unknown, we provide another HMM to estimate ROH in the first step for a sample of sufficient coverage ($\geq 0.5x$). In a second step, the relatedness-HMM uses the contamination-corrected differences between a pair of individuals, and the positions of ROH tracts, to infer relatedness. Our method is available on \url{https://github.com/DivyaratanPopli/KIn_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Relatedness Model overview} 
To estimate the relatedness of two individuals, we subdivide the genome into $L$ large windows $w$ (typically of size 10Mb). The inputs of our algorithm are i) the number of sites $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) probability of ROH in in windows, by default obtained from ROH-HMM described in a following section. Calculation of $N_w$ and $D_w$ requires comparison of reads at all positions for a pair of individuals. For low coverage genomes, this can be done by comparing randomly sampled read at each position for a pair of individuals (or by comparing pseudo haploid sequences). However, we realized that such random sampling can throw away a lot of useful data. To avoid this wastage, our input generation pipeline for Relatedness HMM calculates proportion of reads with alternate allele, at all positions for each genome. We use these proportions to find pairwise heterozygosity as the probability of comparing different alleles. These probabilities are summed up in each window to give $D_w$, along with the number of overlapping sites $N_w$.

Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

The model then uses this data for a given pair of individuals to classify each window into three hidden states $Z_w \in$ ($i_0$, $i_1$, $i_2$), reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($\omega_0$, $\omega_1$, $\omega_2$), reflecting, neither, one or both individuals being in a run-of-homozygosity at this genomic location.

Ideally, we would like to compute the likelihood $P(\BD | \BZ, \BH, \theta)$, where $\theta$ is a vector of parameters (initial transition probability $\pi$, transition matrix $A$ and $N_w$).

However, calculating this likelihood is hard because the number of possible states for $\mathbf{Z}$ is very large, and we use a standard trick for fitting HMMs, to base or inference on the complete data likelihood:

\begin{align}
P(\mathbf{D},\BZ, \BH|\theta) &= P(\mathbf{D}|\mathbf{Z},\BH, \theta) P(\BZ |\theta)P(\BH |\theta) \nonumber\\
&= \prod_w P(D_w|Z_w,H_w, \theta) \prod_w P(H_w |\theta) \prod_w P(Z_w |Z_{w-1},\theta) P(Z_0|\theta)
\label{eq:1}
\end{align}

Here, $P(D_w|Z_w,  H_w, \theta)$ is the emission probability for window $w$,  $P(Z_w|Z_{w-1},\theta)$ is the transition probability, and $P(Z_0| \theta)$ is the initial probability $\pi$. $P(H_w |\theta)$ is the output of ROH HMM described in a following section.

Using the complete data likelihood allows us to factor the emission and transition probabilities, and hence we can maximize them independently.
Transition matrix for different cases of relatedness are either derived from theory, or obtained from simulations (see section). Initial transition probability $\pi_i$ is same for all the $i$.

\subsection{Emission probability}
In equation \ref{eq:1} the term $P(D_w | Z_w, H_w, \theta) = P(D_w | Z_w, H_w, N_w)$ is the emission probability, i.e. the probability of our data , given a particular hidden state. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming sites are equally distributed and independent, we could use a  binomial likelihood
$$P(D_w|Z_w, N_w, H_w) \sim \text{Binom}[D_w ;N_w, p(Z_w, H_w)] \text{,}$$

where, $p$  is the expected proportion of differences we would expect for a particular IBD and ROH state.

If the two individuals are unrelated in a particular window and neither is in a ROH region (i.e. $Z_w = i_0$, and $H_w = \omega_0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes.

$p_0$, $p_1$, and $p_2$ are the expected proportion of differences in $i_0$, $i_1$, and $i_2$. $p_0$ is calculated directly from the data by calculating median of differences for all possible pairs of specimens. We calculate $p_1$ as $(3/4)p_1$ and $p_2$ as $p_1/2$. However, we need one more parameter (represented as $p_3$ in the matrix below) in cases where IBD state is $i_1$ and ROH state is $\omega_2$, or IBD state is $i_2$ and ROH state is $\omega_1$ or $\omega_2$. In such cases the expected proportion of differences between the individuals should be 0. However, as mentioned before, we do all our calculations in large windows, and start/end positions of these windows may not coincide with that of ROH tracts. And so, the pairwise differences in this state (called $p_3$ in the table) can lie anywhere between 0 to $p_2$. We found $p_3$ = $p_2$/2 to be a reasonable assumption for expected pairwise proportion of differences in this IBD/ROH state.

Taken together, we can summarize $p$ as follows 
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_1 & p_2 \\
p_0 & p_2 & p_3 \\
p_0 & p_4 & p_3
\end{array}\right]
\end{equation}

The effect of these considerations is that even though we have nine possible combinations of $Z$ and $H$ for each window, there are actually only four different $p$-parameters $p_x$ with $x$ $\in$ (0,1,2,3).

However, we find that the data often has considerably higher variance than would be expected from a binomial distribution, likely because we average over a large number of SNPs with different allele frequencies.

We therefore add an overdispersion parameter $\delta$ that allows for variation in the $p$ between SNPs. This results in the betabinomial likelihood, (we show that such a distribution fits the data well (fig. S\ref{FigS1:binom})).

\begin{equation}\label{eq:3}
P(D_{w}|p_x,\delta_x,N_w) = \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{x}, N_w-D_w+ \delta_{x}(1-p_{x}))}{ B(p_{x}\delta_{x}, (1-p_{x})\delta_{x})}
\end{equation}

\subsubsection{Estimation}

Emission probabilities are estimated with Expectation-Maximization (EM) algorithm. In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate joint probability of the observed data and the IBD states $Z$, given an initial guess of $\delta$. In maximization step we use these joint probabilities to optimize $\delta$, and update emission probabilities. Iterations are run, until the complete data likelihood stops changing.

\paragraph{Initialization}
The value of $\delta$ is unknown to start with, and is randomly assigned, such that the mean of beta distribution and variance is within constraints described in maximization section.

\paragraph{Expectation step}

In this step, using the standard forward-backward algorithm, we calculate the posterior probability of each IBD state in each window ($\gamma_{wi}$). Expectation step, among other parameters, requires emission probabilities. We calculate emission probabilities for each case of $Z_w$ and $H_w$ from $\delta$ and $p$ using equation \ref{eq:3}, and we get the total emission probabilities for each $Z_w$ as $\sum_{H_w} P(D_w|Z_w,N_w,H_w) P(H_w|\theta)$

\paragraph{Maximization step}

The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_j$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.

\begin{equation}\label{eq:4}
\mathcal{C} = \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta, p)h_{w\omega}\gamma_{wi}
\end{equation}

@@

\begin{align*}
k_{w0} &= \gamma_{w0}\\
k_{w1} &= \gamma_{w1} h_{w0}\\
k_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
k_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite equation \ref{eq:4} as:
\begin{align}
\mathcal{C} &= \sum_{w=1}^L\sum_{x=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_x, p_x)k_{wx}
\end{align}

>>>>

\documentclass[12pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[super]{nth}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage[
singlelinecheck=false
]{caption}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{bbm}
\usepackage{graphicx}
\usepackage{csvsimple}
\usepackage[section]{placeins}

@@

\newcommand{\BZ}{\mathbf{Z}}
\newcommand{\BD}{\mathbf{D}}
\newcommand{\BN}{\mathbf{N}}
\newcommand{\BH}{\mathbf{H}}
\newcommand{\Btheta}{\pmb{\theta}}

@@

\noindent Knowledge of familial relationships is an important part of several research fields. However, estimating relatedness from ancient DNA can be be difficult since DNA sequences extracted from ancient bones may have low coverage, ascertainment bias (if the sequences have been captured with DNA probes), and contamination from modern humans. In addition, the population of interest may have long runs of homozygosity (ROH) due to recent inbreeding or/and small population size. This can affect the relatedness estimates. Here, we present a Hidden Markov Model (HMM) to estimate identity by descent (IBD) fragments and degree of relatedness up to 3rd degree, while additionally differentiating between siblings and parent-child. The output can be further refined with another HMM that we developed to estimate ROH proportions.
\end{abstract}

@@

Identifying related individuals is a common task in applications of genetics. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify missing persons following a disaster \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential pre-processing step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an  understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals in the mating animals is important to ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

These analyses use cranial and dental metric and morphological and postcranial non-metric data to infer inter-individual relatedness based on the assumption that the phenotype reflects underlying genetic variation and patterns of relatedness. Multifactorial inheritance complicates straightforward interpretations, however, and given the moderate to low (and population specific) heritabilities of most morphometric traits (see Scott & Turner, 1997 for an overview; also Townsend et al., 2009), the effects of development on final phenotypic form are difficult to specify.

@@

Commonly, pairs of related individuals are  identified by looKINg for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}.  One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999,mcpeek_statistical_2000}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from average of observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, maKINg it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ. This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when maKINg pedigrees.

\subsection{How our method works}
Here, we present KIN (KINship Inference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment bias. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

@@

The core of  KIN is a HMM that aims to infer relatedness and IBD sharing between a pair of low-coverage individuals, optionally taking ROH tracts and contamination estimates in each individual into account. If the locations of ROH tracts are unknown, we provide another HMM to coarsely estimate the location of ROH for samples  of sufficient coverage ($\geq 0.5x$). Our method is available on \url{https://github.com/DivyaratanPopli/KIN_snakemake} along with a \textt{snakemake} \cite{koster_snakemakescalable_2012} pipeline to generate the input files for the models directly from bam files.

\subsection{Relatedness Model overview} 
Our relatedness model assumes we have low-coverage genome sequences for a number of individuals from the same population, and our goal is to infer which pairs of individuals are related. We subdivide the genome into $L$ large windows (typically of size 10Mb). The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in a following section. $D_w$ is the number of sites that are different between a pair of individuals, and can be estimated accurately for high coverage data by comparing genotypes. For low coverage samples, $D_w$ is typically calculated by comparing randomly-sampled reads (or pseudo-haploid sequences) \cite{haak_massive_2015}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_s \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the derived allele frequencies for individuals i and j respectively, at SNP index $s$ in a window.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$. The KIN-HMM uses $\BD$ and $\BN$  to classify each window into three hidden states $Z_w \in$ ($0$, $1$, $2$), reflecting zero, one or two chromosomes shared IBD. Likewise, we classify each window into one of three possible ROH state $H_w \in$ ($0$, $1$, $2$), reflecting, zero, one or both individuals being in a run-of-homozygosity at this genomic location.

There are two additional parameters, $\pi$ and $\mathbf{A}$. The initial probabilities $\pi$ give the probabilities of being in each state $Z$ at the beginning of each chromosome, which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed for each relatedness category (see section YYY). For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

We compute the complete data likelihood for the HMM:

\begin{align}\label{eq:1}
\log P(\BD,\BZ|\BH, \theta) &= \log P(\BD|\BZ,\BH, \theta) + \log P(\BZ |\theta) \nonumber\\
&= \sum_w \log P(D_w|Z_w,H_w, \theta) + \sum_w \log P(Z_w |Z_{w-1},\theta) + \log P(Z_0|\theta)
\end{align}

\subsection{Emission probability}
Using this set up, we can isolate the emissions $P(D_w | Z_w, H_w ,\theta)$ from equation \ref{eq:1} and optimize them alone. The number of shared sites $N_w$ is the only parameter in $\theta$ that affects the emissions.

Assuming sites are equally distributed and independent would result in the binomial likelihood:
$$P(D_w|Z_w, H_w, N_w) \sim \text{Binom}[D_w ; p(Z_w, H_w), N_w] \text{,}$$

where, $p$ is the expected proportion of differences we would expect for a particular IBD and ROH state. If the two individuals are unrelated in a particular window (i.e. $Z_w = 0$), then the expected proportion of pairwise differences depends solely on the effective population size, and we denote this proportion with $p_0$. Note that $p_0$ does not depend on the presence of ROH, since this does not change expected proportion of differences between unrelated individuals. If the two individuals share one or even both copies of the genome IBD, we would expect the proportion of differences to be reduced to $p_1 = \frac{3}{4} p_0$, and $p_2 = \frac{1}2 p_0$, respectively, since either one or two of the four possible comparisons will be between identical chromosomes \cite{kuhn_estimating_2018}. Thus, $p(Z_w=i, H_w=0) = p_i$.

The proportion of differences between unrelated individuals $p_0$ is an important parameter. We follow READ \cite{kuhn_estimating_2018} and estimate it as the median of differences for all possible pairs of individuals, which should work well if the majority of individuals in the samples are unrelated. We set $p_1$ to $(3/4)p_0$ and $p_2$ to $p_0/2$.

The presence of inbreeding adds an additional complication, as it is possible that all chromosomes are shared between individuals, so that the expected differences should be zero. An example would be inbred regions in two bones from the same individual. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that setting expected proportion of differences in this case ($p_3$) at $p_2/2$ works well in practice.

Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_3 \\
p_2 & p_3 & p_3
\end{array}\right]
\end{equation}

The effect of these considerations is that even though we have nine possible combinations of $Z_w$ and $H_w$ for each window, there are actually only four unique $p$-parameters $p_i$ with $i$ $\in$ (0,1,2,3).

\paragraph{Beta Binomial Model}
The Binomial Model laid out above assumes that sites are independent. However, due to genetic linkage, nearby sites will be correlated and we find that the data often has considerably higher variance than would be expected from a binomial model. We take this into account by adding an overdispersion parameter $\delta$ that allows for variation in the $p$ between windows. Note that, just like $p(Z_w,H_w)$, $\delta(Z_w,H_w)$ has effectively four unique parameters $\delta_i$ with $i$ $\in$ (0,1,2,3). This is due to the fact that several combinations of $Z_w$ and $H_w$ can lead to comparison between same number of unique chromosomes, and such states will have same mean as well as the same variance. This results in the beta-binomial likelihood, that can be fitted to the data (fig. S\ref{figS1:binom})).

\begin{align}\label{eq:3}
P(D_{w}|p_i,\delta_i,N_w) &\sim BB[D_w; N_w, p_i, \delta_i] \nonumber\\
&= \binom{N_w}{D_w}\frac{B(D_w+p_x \delta_{i}, N_w-D_w+ \delta_{i}(1-p_{i}))}{ B(p_{i}\delta_{i}, (1-p_{i})\delta_{i})}.
\end{align}

This parameterization of the beta distribution in terms of expected value $p$ and overdispersion $\delta$ is also called the Balding-Nichols-model \cite{balding_method_nodate}, and is distinct from the more common parametrization in terms of $\alpha$ and $\beta$. It is worth mentioning here that $D_w$ is estimated as a sum of probabilities (see section XX), while both $D_w$ and $N_w$ may undergo contamination correction, which causes these estimates to be non-integral. We calculate factorials required in Beta Binomial likelihood with Gamma function, which is continuous and hence does not require integral values.

\subsection{Estimation}

Emission probabilities are estimated with Expectation-Maximization (EM) algorithm. In each iteration of the algorithm, the Expectation step involves forward-backward algorithm to calculate $P(D, Z| H, \theta, \delta)$.

In maximization step we use these joint probabilities to optimize $\delta_i$ for each state i $\in$ (0,1,2,3), and update emission probabilities. Iterations are run, until the complete data log likelihood stops changing.

\paragraph{Initialization}
The value of $\delta_i$ is unknown to start with, and is set to a random value.

\paragraph{Expectation step}
In the $t$-th iteration, we calculate the posterior probability of each IBD state in each window $\gamma^{(t)}_{wi} = P(Z_w=i | \BD, \BH, \theta, \delta^{(t)})$ using the forward-backward algorithm, where $\delta^{(t)}$ is the current guess for delta.

\paragraph{Maximization step}
The only free parameters we estimate in the M-step are the overdispersion parameters $\delta_i$. We do this optimization using a cost function, which is the log-emission probability weighted by the posterior probabilities of the hidden states $\gamma_{wi}$ and the ROH state-probabilities $h_{w\omega}$.

\begin{align}\label{eq:4}
\mathcal{C} &= \mathbb{E}[\log P(D_w|Z_w, H_w, \theta, \delta^{(t-1)}]\nonumber\\
&= \sum_{w=1}^L \sum_{i=0}^2\sum^2_{\omega=0} \log P(D_{w}|Z_w, H_w, N_w, \delta^{(t-1)}, p)h_{w\omega}\gamma_{wi}
\end{align}

@@

\begin{align*}\label{eq:5}
g_{w0} &= \gamma_{w0}\\
g_{w1} &= \gamma_{w1} h_{w0}\\
g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{w=1}^L\sum_{i=0}^3 \log P(D_{w}|Z_w, H_w, N_w, \delta_i, p_i)g_{wi}
\end{align}

Since each term in equation \ref{eq:6}  depends on only one $\delta_i$ and not on different IBD and ROH states, we can optimize for them independently.
